Sample Length at Which Block Bootstrapping is Effective for Estimation of Parameter of Time Series

### 2023-01-17 Manuscript/manuscript.tex
Clean Manuscript Folder

Block bootstrapping is a method that can be applied to a non-stationary process to estimate a parameter. Suppose that a sample of length n of a time series is given. In regular bootstrap procedure, the new pseudo-sample of the same size as the original sample would be created by simply resampling observations from the with replacement. In the case of a time series, in order to account for the temporal dependence, the time series can be split into blocks, typically of the same size. The block should be of size l large enough to include time dependence, yet small enough to include some variance. l should also divide n. As n increases, an ideal l should also increase, but the ratio of l to n should decrease. To achieve this, l is often assigned a value as a function of n. A common function that is considered the best by much previous literature is l = n$^{1/3}$. To make things sample for this study, only perfect cubes were used for the sample length. A bootstrapped sample is created by taking a sample of n / l blocks with replacement to create a new pseudo-sample of size n. The designer of the study can choose whether the blocks overlap or not. If the blocks can never overlap, it is called a non-moving block bootstrap. In this case, the original sample is split evenly into n / l blocks, and the same number of blocks are sampled with replacement to create a bootstrapped pseudo-sample. If the blocks do overlap, it is called a moving block bootstrap. n / l samples are taken randomly from the original sample, but they do not have to be from a set of evenly spaced blocks. An estimate of the parameter (or parameters) is computed from the bootstrapped sample. This procedure can be repeated many times to create a sampling distribution of the estimator. In this study, 1000 bootstrapped pseudo-samples were created for each simulation. Using this distribution of point estimates, a 1 - $\alpha$/2 \% confidence interval for the parameter can be created. The most simplistic method for constructing this interval is the percentile confidence interval. For a parameter $\theta$, the percentile confidence interval takes the form: \[ [ ]\]. Another method for constructing this interval is the empirical confidence interval, which takes the form: \[ [ ]\]. The last method used in this study is the bias-corrected and accelerated 
(BCA) confidence interval, which takes the form: \[ [ ]\].

@@

The central objective of the study was to assess what sample size is necessary for the the block bootstrap method to estimate a parameter of an autoregressive process. The method will work better as the sample size goes to infinity, but the goal is to see what is the smallest sample size that is acceptable. The sample size is acceptable if many 1 - $\alpha$/2 confidence intervals for that sample size recover the parameter at a rate of 1 - $\alpha$/2. In the previous section, the procedure for creating a confidence interval of a parameter using the block bootstrap method is described. This process is applied to a simulation of an autoregressive integrated moving average (ARIMA) process to create a confidence interval of a certain parameter (could be the mean or the AR(1) coefficient). R has a built in ARIMA simulation which allows you to set the mean, AR(1) coefficient, and the standard deviation of the error term. Using the equation below, the standard deviation of the error term can be set so that the standard deviation of Y is constant (0.1, Variance = 0.01).

\[ \sigma_{x}^{2}=\sigma_{\eta}^{2}/\left( 1-\phi^2 \right)\]
\[\sigma_{\eta}=\sqrt{\sigma_{x}^{2}*\left( 1-\phi^2 \right)}\]

For all simulation studies, the mean can be kept at 0, but the AR(1) Coefficient is the main factor that makes block bootstrap less effective, so its impact on the method's performance should be analyzed. For a certain variation of the ARIMA process, block bootstrap 95\% confidence intervals for each of 3 parameters (mean, AR(1) Coefficient, and sigma-squared) for a sample of length 32 is replicated 4000 times (each replication of the method is applied to a newly generated ARIMA simulation). For each individual confidence interval, it can be recorded whether or not the interval includes the true parameter. The coverage rate, or the proportion of intervals that include the true parameter, can then be recorded. If the parameter and its variance are being properly estimated by the block bootstrap, the coverage rate should reflect this by being close to the confidence level of 95\%. If this is not the case, the method is not working as well as it should. The width of the confidence interval for each replication is dependent on the variance of the parameter of the time series. If the interval is on average not wide enough to capture the parameter at the expected rate of 95\%, it indicates that the variance of the parameter (which will be estimated from the standard error of the parameter estimates from the bootstrapped samples) will be underestimated. If the interval is too wide and captures the parameter at a rate above 95\%, it indicates that the variance of the parameter will be overestimated. Since the coverage rate is a proportion based on many replication of independent Bernoulli outcomes, a 95\% confidence interval of the coverage rate at that sample size can also be created. We can be 95\% confident that the true expected coverage rate is included in this interval, so if .95 is included in the interval, this indicates that the block bootstrap method is working well. Once the simulation is done for this sample size, the process can be run again with a sample twice as big. This can be continued until the method is determined to be working acceptably (.95 is included within the confidence interval of the coverage rate).

@@

6 variations of simulation studies were analyzed based on factors that might affect the results of the block bootstrap method. Simulation studies were conducted for an autoregressive process with AR = 0, an autoregressive process with AR = 0.2, and an autoregressive process with AR = 0.4. For each of these AR(1) coefficient values, separate simulation studies were conducted for the mean and AR(1) coefficient. For each of these variations, the non-moving block bootstrap method as well as the moving block bootstrap method were applied to estimate the respective parameter.

The goal was to determine what range of sample sizes was acceptable for each of these variations of simulation studies. For theta = 0, results for five sample sizes (32, 64, 128, 256, 512) were observed. For theta = 0.2, results for eight sample sizes (32,  64, 128, 256, 512, 1024, 2048, 4096) were observed. For theta = 0.4, results for nine sample sizes (32, 64, 128, 256, 512, 1024, 2048, 4096, 8192) were observed.

@@

As the AR(1) coefficient increases, it was found unsurprisingly that a large sample size was needed for the block bootstrap method to work. When the target for estimation was the mean or the variance of the process, moving block bootstrap generally had a similar performance to non-moving block bootstrap. However when the target for estimation was the AR(1) coefficient of the process, moving block bootstrap noticeably recovered the parameter at a greater rate than non-moving block bootstrap. When the target was the mean or the variance and theta was more than 0, the coverage rate was generally much higher than that for when the target was the AR(1) coefficient. However when theta was 0, the AR(1) coefficient was recovered at a greater rate than the mean or variance .

>>>>

Block bootstrapping is a method that can be applied to a non-stationary process to estimate a parameter. Suppose that a sample of length n of a time series is given. In regular bootstrap procedure, the new pseudo-sample of the same size as the original sample would be created by simply resampling observations from the with replacement. In the case of a time series, in order to account for the temporal dependence, the time series can be split into blocks, typically of the same size. The block should be of size l large enough to include time dependence, yet small enough to include some variance. l should also divide n. As n increases, an ideal l should also increase, but the ratio of l to n should decrease. To achieve this, l is often assigned a value as a function of n. A common function that is considered the best by much previous literature is l = n$^{1/3}$. To make things sample for this study, only perfect cubes were used for the sample length. A bootstrapped sample is created by taking a sample of n / l blocks with replacement to create a new pseudo-sample of size n. The designer of the study can choose whether the blocks overlap or not. If the blocks can never overlap, it is called a non-moving block bootstrap. In this case, the original sample is split evenly into n / l blocks, and the same number of blocks are sampled with replacement to create a bootstrapped pseudo-sample. If the blocks do overlap, it is called a moving block bootstrap. n / l samples are taken randomly from the original sample, but they do not have to be from a set of evenly spaced blocks. An estimate of the parameter (or parameters) $\hat{\theta}_{b}$ is computed from the bootstrapped sample. This procedure can be repeated many times to create a distribution of $\hat{\theta}_{b}$. In this study, 1000 bootstrapped pseudo-samples were created for each simulation. Using this distribution of $\hat{\theta}_{b}$, a 1 - $\alpha$/2 \% confidence interval for the parameter can be created. The most simplistic method for constructing this interval is the percentile confidence interval. For a parameter $\theta$, the percentile confidence interval takes the form: \[ [\hat{\theta}_{b, .025}, \hat{\theta}_{b, .975}].\] Another method is the bias-corrected and accelerated 
(BCA) confidence interval, which takes the form: \[ [\hat{\theta}_{b, \alpha_{L}}, \hat{\theta}_{b, \alpha_{U}}].\] $\alpha{L}$ and $\alpha{U}$ are the cumulative probability of $z{L}$ and $z{U}$, respectively, where:
\[z_{L} = \frac{z_{0} - z_{1 - \alpha/2}}{1 - a(z_{0} - z_{1 - \alpha/2})} + z_{0}\] and
\[z_{U} = \frac{z_{0} + z_{1 - \alpha/2}}{1 - a(z_{0} + z_{1 - \alpha/2})} + z_{0}.\] 
	Both of these intervals are good at recovering the mean and standard deviation of a temporally dependent process, but their coverage of the temporal dependence deteriorates as n increases. A potential solution to this issue is to center the intervals around either the original sample estimate $\hat{\theta}$ or the mean of all $\hat{\theta}_{b}$.

@@

The central objective of the study was to assess what sample size is necessary for the the block bootstrap method to estimate a parameter of an autoregressive process. The method will work better as the sample size goes to infinity, but the goal is to see what is the smallest sample size that is acceptable. The sample size is acceptable if many 1 - $\alpha$/2 confidence intervals for that sample size recover the parameter at a rate of 1 - $\alpha$/2 \%. In the previous section, the procedures for creating three types of confidence interval of a parameter using the block bootstrap method are described.

%These procedures are applied to a simulation of an autoregressive integrated moving average (ARIMA) process to create three confidence interval of certain parameters (the mean, the AR(1) coefficient, the MA(1) coefficient, the standard deviation, etc.).

R has a built in ARIMA simulation function which allows you to set the mean, AR and MA coefficients, and the standard deviation of the error term $\epsilon$. The standard deviation $\sigma_{\epsilon}$ of a AR(1) process with mean 0 can be written as a function of $\sigma_{x}$ and the AR(1) coefficient $\theta$. 
\[X_{t} = \epsilon_{t} + \theta*X_{t-1}\]
\[Var(X_{t}) = Var(\epsilon_{t} + \theta*X_{t-1})\]
\[\sigma^2_{x} = \sigma^2_{\epsilon} + \theta^2*\sigma^2_{x}\]
\[\sigma^2_{x}(1 - \theta^2) = \sigma^2_{\epsilon}\]
\[\sigma_{\epsilon} = \sigma_{x}\sqrt{1 - \theta^2}\]

For the purposes of this study, $\mu$ should be fixed at 0, and $\sigma_{x}$ at 1. By default, $\mu$ is 0 and $\sigma_{\epsilon}$ is 1, so the observations can be multiplied by $\sqrt{1 - \theta^2}$ so that the theoretical $\sigma_{x}$ is equivalent to 1.

For all simulation studies, , but $\mu$ and $\sigma_{x}$ can be kept constant, but increasing temporal dependence factors such as AR and MA Coefficients can make block bootstrap less effective in estimation, so its impact on the method's performance should be analyzed. For a certain variation of an AR(1) process, block bootstrap 95\% confidence intervals for each of 3 parameters ($\mu$, $\sigma_{x}$, and the autocorrelation function$\rho$) can be replicated for 1000 simulated samples of a certain length. For each individual confidence interval, it can be recorded whether or not the interval includes the true parameter. The coverage rate, or the proportion of intervals that include the true parameter, can then be recorded. If the parameter is being properly estimated by the block bootstrap, the coverage rate should reflect this by being close to the confidence level of 95\%. If this is not the case, the method is not working as well as it should. The width of the confidence interval for each replication is dependent on the variability of the parameter of the time series. If the interval is on average not wide enough to capture the parameter at the expected rate of 95\%, it indicates that the variability of the parameter is underestimated. If the interval is too wide and captures the parameter at a rate above 95\%, it indicates that the variability of the parameter is overestimated. Since the coverage rate is a proportion based on 1000 independent binary outcomes (the interval recovers the parameter or it doesn't), a 95\% confidence interval of the coverage rate for each parameter can also be created. The inclusion of .95 in the interval indicates that the block bootstrap method is working well for its respective parameter. If this is not the case for this sample length, the procedure can be conducted again with a larger sample length. This can be repeated until the method is determined to be working acceptably, meaning that .95 is included in the 95\% confidence interval of the coverage rate of the parameter.

@@

1000 replicates of length-n random samples of an AR(1) process with theoretical $\mu$ = 0, $\sigma_{x}$ = 1, and $\theta$ = 0.1 were simulated using R. For each of three target parameters ($\mu$ = 0, $\sigma_{x}$ = 1, and $\theta$ = 0.1, three types of block bootstrap (with block length l) confidence intervals (percentile, empirical, and BCA) were constructed for each replicate sample. The statistic used to estimate $\mu$ is the sample mean $\bar{x}$, the statistic used to estimate $\sigma$ is the sample standard deviation $S^2$, and the statistic used to estimate $\theta$ is the auto-correlation function of the sample. For each of these nine categories of confidence intervals, the coverage rate of its respective parameter was computed, along with a 95\% confidence interval of the proportion given 4000 trials. For a certain variation of the process, the study was repeated for n = 125, 1000, 3375, and 8000. The only parameter of the process that was varied was $\theta$, and 0.1, 0.3, and 0.5 were the values used.

@@
